---
title: "Final_Project_Code"
output: html_document
---
```{r setup}
 #knitr::opts_knit$set(root.dir = normalizePath("/home1/gittings/490_cluster/analysis_data")) #      knitr::opts_knit$set(root.dir = normalizePath("/home1/emmashih/490_cluster/analysis_data")) 
      knitr::opts_knit$set(root.dir = normalizePath("/home1/spoden/490_cluster/analysis_data")) 
```

```{r}
library(TCGAbiolinks)
library(maftools)
library(ggplot2)
```

```{r}
clin_query <- GDCquery(project = "TCGA-SKCM",
data.category = "Clinical",
data.type = "Clinical Supplement",
data.format = 'SKCM Biotab')
# GDCdownload(clin_query)
clinical.BCRtab.all <- GDCprepare(clin_query)
clinic <- clinical.BCRtab.all$clinical_patient_skcm[-c(1,2),]
```

```{r Exercise 4.2}
clinicSKCM$stage_simple <- ifelse(grepl("^Stage IV", clinicSKCM$ajcc_pathologic_tumor_stage), 4,
                           ifelse(grepl("^Stage III", clinicSKCM$ajcc_pathologic_tumor_stage), 3,
                           ifelse(grepl("^Stage II", clinicSKCM$ajcc_pathologic_tumor_stage), 2,
                           ifelse(grepl("^Stage I($|[^VIX])", clinicSKCM$ajcc_pathologic_tumor_stage), 1, NA))))

clinicSKCM$age <- as.numeric(clinicSKCM$age_at_diagnosis)

clinicSKCM_scatter <- clinicSKCM[!is.na(clinicSKCM$stage_simple) & !is.na(clinicSKCM$age), ]

plot(jitter(clinicSKCM_scatter$stage_simple, amount = 0.1), clinicSKCM_scatter$age,
     xlab = "Tumor Stage",
     ylab = "Age at Diagnosis",
     main = "Scatterplot of Age vs Tumor Stage",
     pch = 19)
```{r}
ggplot(clinic, aes(x = ajcc_pathologic_tumor_stage)) + geom_bar()
```

```{r}
library(dplyr)
library(survival)
library(survminer)

colnames(clinic)
```
```{r}

clinic <- clinic %>%
  mutate(
    death_days_to        = as.numeric(death_days_to),
    last_contact_days_to = as.numeric(last_contact_days_to),
    age_at_diagnosis     = as.numeric(age_at_diagnosis),
    breslow_thickness_at_diagnosis = as.numeric(breslow_thickness_at_diagnosis),
    
    # Unified survival time: use death days if dead, otherwise last contact
    os_time  = ifelse(!is.na(death_days_to), death_days_to, last_contact_days_to),
    os_event = ifelse(vital_status == "Dead", 1, 0)
  )
```
```{r}
fit <- survfit(Surv(os_time, os_event) ~ 1, data = clinic)

ggsurvplot(fit, data = clinic,
           #risk.table = TRUE,
           conf.int   = TRUE,
           title      = "TCGA SKCM: Overall Survival",
           xlab       = "Days",
           ylab       = "Survival Probability")
```


Stacked Bar Plot 
```{r}

```

random plot

```{r}
library(TCGAbiolinks)
library(maftools)
```


boxplot of Breslow thickness across AJCC tumor stages (the one from the exit ticket but better verison)
```{r}
library(ggplot2)
library(dplyr)

# Clean the stage and Breslow columns
breslow_clean <- clinic %>%
  filter(
    !is.na(ajcc_pathologic_tumor_stage),
    !is.na(breslow_thickness_at_diagnosis),
    breslow_thickness_at_diagnosis > 0,
    !ajcc_pathologic_tumor_stage %in% c("[Not Available]", "[Discrepancy]", 
                                         "[Unknown]", ""),
    !breslow_thickness_at_diagnosis %in% c("[Not Available]", "[Unknown]", "")
  ) %>%
  mutate(
    breslow_thickness_at_diagnosis = as.numeric(breslow_thickness_at_diagnosis),
    # Simplify stage to I, II, III, IV (strip substage letters like IA, IIB, etc.)
    stage_simple = gsub("Stage ", "", ajcc_pathologic_tumor_stage),
    stage_simple = gsub("[ABC]$", "", stage_simple),
    stage_simple = factor(stage_simple, levels = c("I", "II", "III", "IV"))
  ) %>%
  filter(!is.na(stage_simple), !is.na(breslow_thickness_at_diagnosis))

ggplot(breslow_clean, aes(x = stage_simple, y = breslow_thickness_at_diagnosis, 
                           fill = stage_simple)) +
  geom_boxplot(outlier.shape = 21, outlier.alpha = 0.5) +
  scale_fill_manual(values = c("I" = "blue", "II" = "red", 
                                "III" = "pink", "IV" = "green")) +
  scale_y_log10() +   # log scale because Breslow thickness is right-skewed
  xlab("AJCC Tumor Stage") +
  ylab("Breslow Thickness at Diagnosis (mm, log scale)") +
  ggtitle("Tumor Invasion Depth Across SKCM Stages")
```

Patient count by stage and vital status
```{r}
library(ggplot2)
library(dplyr)

stage_survival <- clinic %>%
  filter(
    !is.na(ajcc_pathologic_tumor_stage),
    !ajcc_pathologic_tumor_stage %in% c("[Not Available]", "[Discrepancy]", 
                                         "[Unknown]", ""),
    !is.na(vital_status),
    !vital_status %in% c("[Not Available]", "[Unknown]", "")
  ) %>%
  mutate(
    stage_simple = gsub("Stage ", "", ajcc_pathologic_tumor_stage),
    stage_simple = gsub("[ABC]$", "", stage_simple),
    stage_simple = factor(stage_simple, levels = c("I", "II", "III", "IV")),
    vital_status = factor(vital_status, levels = c("Alive", "Dead"))
  ) %>%
  filter(!is.na(stage_simple))

axis(1, at = c(1, 2, 3, 4), labels = c("Stage I", "Stage II", "Stage III", "Stage IV"))
```






