---
title: "Final_Project_Code"
output: html_document
---
```{r setup}
 #knitr::opts_knit$set(root.dir = normalizePath("/home1/gittings/490_cluster/analysis_data")) 
#knitr::opts_knit$set(root.dir = normalizePath("/home1/emmashih/490_cluster/analysis_data")) 
       knitr::opts_knit$set(root.dir = normalizePath("/home1/spoden/490_cluster/analysis_data")) 
```

```{r}
library(TCGAbiolinks)
library(maftools)
library(ggplot2)
```

```{r}
# code to download and load CLINICAL DATA
clin_query <- GDCquery(project = "TCGA-SKCM",
data.category = "Clinical",
data.type = "Clinical Supplement",
data.format = 'SKCM Biotab')
# GDCdownload(clin_query)
clinical.BCRtab.all <- GDCprepare(clin_query)
clinic <- clinical.BCRtab.all$clinical_patient_skcm[-c(1,2),]
```



```{r}
library(dplyr)
library(survival)
library(survminer)
```
```{r}

clinic <- clinic %>%
  mutate(
    death_days_to        = as.numeric(death_days_to),
    last_contact_days_to = as.numeric(last_contact_days_to),
    age_at_diagnosis     = as.numeric(age_at_diagnosis),
    breslow_thickness_at_diagnosis = as.numeric(breslow_thickness_at_diagnosis),
    
    # Unified survival time: use death days if dead, otherwise last contact
    os_time  = ifelse(!is.na(death_days_to), death_days_to, last_contact_days_to),
    os_event = ifelse(vital_status == "Dead", 1, 0)
  )
```
```{r}
fit <- survfit(Surv(os_time, os_event) ~ 1, data = clinic)

ggsurvplot(fit, data = clinic,
           #risk.table = TRUE,
           conf.int   = TRUE,
           title      = "TCGA SKCM: Overall Survival",
           xlab       = "Days",
           ylab       = "Survival Probability")
```

Patient count by stage and vital status

```{r}
library(ggplot2)
library(dplyr)
count  <- dplyr::count
select <- dplyr::select
filter <- dplyr::filter

stage_survival <- clinic %>%
  filter(
    !is.na(ajcc_pathologic_tumor_stage),
    !ajcc_pathologic_tumor_stage %in% c("[Not Available]", "[Discrepancy]", 
                                         "[Unknown]", ""),
    !is.na(vital_status),
    !vital_status %in% c("[Not Available]", "[Unknown]", "")
  ) %>%
  mutate(
    stage_simple = gsub("Stage ", "", ajcc_pathologic_tumor_stage),
    stage_simple = gsub("[ABC]$", "", stage_simple),
    stage_simple = trimws(stage_simple),
    stage_simple = factor(stage_simple, levels = c("I", "II", "III", "IV")),
    vital_status = factor(vital_status, levels = c("Alive", "Dead"))
  ) %>%
  filter(!is.na(stage_simple))

stage_pct <- stage_survival %>%
  count(stage_simple, vital_status) %>%
  group_by(stage_simple) %>%
  mutate(
    pct = n / sum(n) * 100,
    label = paste0(round(pct, 1), "%")
  )

ggplot(stage_pct, aes(x = stage_simple, y = n, fill = vital_status)) +
  geom_bar(stat = "identity") +
  geom_text(
    aes(label = label),
    position = position_stack(vjust = 0.5),
    size = 3.5,
    color = "black",
    fontface = "bold"
  ) +
  scale_fill_manual(values = c("Alive" = "pink", "Dead" = "seagreen"),
                    name = "Vital Status") +
  labs(
    title = "SKCM Patient Count and Survival Outcome by Tumor Stage",
    x = "AJCC Tumor Stage",
    y = "Number of Patients"
  ) +
  theme_gray()
```


Ttranscriptomic volcano plot stages 2 to 3

```{r} 
library(TCGAbiolinks)
library(limma)
library(ggplot2)
library(SummarizedExperiment)
library(dplyr)

# ── 1. Download & prepare (same as before) ────────────────────────────────────
query_rna <- GDCquery(
  project = "TCGA-SKCM",
  data.category = "Transcriptome Profiling",
  data.type = "Gene Expression Quantification",
  workflow.type = "STAR - Counts"
)

#GDCdownload(query_rna)
rna_se <- GDCprepare(query_rna)

# ── 2. Extract counts + metadata ──────────────────────────────────────────────
counts_mat <- assay(rna_se, "unstranded")

sample_meta <- as.data.frame(colData(rna_se)) %>%
  mutate(
    barcode = rownames(.),
    stage_simple = gsub("Stage ", "", ajcc_pathologic_stage),
    stage_simple = gsub("[ABC]$", "", trimws(stage_simple))
  ) %>%
  filter(stage_simple %in% c("II", "III"),
         !is.na(stage_simple))

counts_sub <- counts_mat[, sample_meta$barcode]

```

```{r}
# ── 3. Filter lowly expressed genes + log2 transform ─────────────────────────
keep <- rowSums(counts_sub >= 10) >= 5
counts_sub <- counts_sub[keep, ]

mval <- t(apply(counts_sub, 1, function(x) log2(x + 1)))

# ── 4. Fit limma model (Stage III vs Stage II) ────────────────────────────────
sample_meta$is_stageIII <- sample_meta$stage_simple == "III"

design <- model.matrix(~ is_stageIII, data = sample_meta)
fit    <- lmFit(mval, design)
fit2   <- eBayes(fit)

# ── 5. Build results dataframe ────────────────────────────────────────────────
dat <- data.frame(
  foldchange = fit[["coefficients"]][, 2],
  logPvalue  = -log10(p.adjust(fit2[["p.value"]][, 2], method = "BH")),
  geneName   = rownames(counts_sub)
)

dat$threshold <- as.factor(abs(dat$foldchange) < 1)

#name shit
BiocManager::install("biomaRt")
library(biomaRt)

# Connect to Ensembl
mart <- useMart("ensembl", dataset = "hsapiens_gene_ensembl")

# Strip the version number (the .22 part)
ensembl_ids <- gsub("\\..*", "", dat$geneName)

# Query for gene symbols
gene_map <- getBM(
  attributes = c("ensembl_gene_id", "hgnc_symbol"),
  filters    = "ensembl_gene_id",
  values     = ensembl_ids,
  mart       = mart
)

# Map symbols back to dat
dat$ensembl_clean <- gsub("\\..*", "", dat$geneName)
dat <- left_join(dat, gene_map, by = c("ensembl_clean" = "ensembl_gene_id")) %>%
  mutate(geneName = ifelse(hgnc_symbol == "" | is.na(hgnc_symbol), 
                           ensembl_clean, hgnc_symbol))

# ── 6. Volcano plot (same style as your homework) ─────────────────────────────
library(ggrepel)

# Add direction column
dat <- dat %>%
  mutate(direction = case_when(
    foldchange >  1 & logPvalue > 2 ~ "Upregulated in III",
    foldchange < -1 & logPvalue > 2 ~ "Downregulated in III",
    TRUE ~ "Not significant"
  ))

sig_genes <- dat %>% filter(direction != "Not significant")

# Force label specific genes of interest
force_label <- dat %>% 
  filter(geneName %in% c("FOXQ1", "TP53AIP1", "KRT14", "C7"))

cols <- c("Upregulated in III"   = "firebrick",
          "Downregulated in III" = "steelblue",
          "Not significant"      = "grey70")

ggplot(data = dat, aes(x = foldchange, y = logPvalue)) +
  # Grey background dots first
  geom_point(data = dat %>% filter(direction == "Not significant"),
             aes(colour = direction), alpha = 0.2, size = 0.6) +
  # Significant genes on top
  geom_point(data = dat %>% filter(direction != "Not significant"),
             aes(colour = direction), alpha = 0.9, size = 1.5) +
  scale_colour_manual(values = cols) +
  # Label only the forced genes
  geom_text_repel(
    data = force_label,
    aes(label = geneName),
    size = 3,
    max.overlaps = 20,
    box.padding = 0.5,
    fontface = "bold"
  ) +
  geom_vline(xintercept =  1, colour = "darkred", linetype = "dashed") +
  geom_vline(xintercept = -1, colour = "darkred", linetype = "dashed") +
  geom_hline(yintercept =  2, colour = "darkred", linetype = "dashed") +
  xlab("Fold Change") +
  ylab("-log10 p value") +
  ggtitle("Volcano Plot: Stage III vs Stage II (TCGA-SKCM)") +
  theme_bw() +
  theme(legend.position = "top")
```
significant genes
```{r}
# Explicitly use dplyr's select
dat %>%
  filter(direction != "Not significant") %>%
  dplyr::select(geneName, foldchange, logPvalue, direction) %>%
  arrange(desc(logPvalue))
```



YES KEEP methylation volcano plot stages 2 to 3
attempted volcano methylation plot

```{r}
library(TCGAbiolinks)
library(limma)
library(ggplot2)
library(ggrepel)
library(dplyr)
```

```{r}
if (!requireNamespace("remotes", quietly = TRUE)) {
  install.packages("remotes")
}
library("remotes")
remotes::install_github("BioinformaticsFMRP/TCGAbiolinks")
library(TCGAbiolinks)
```

```{r}
query_meth <- GDCquery(
  project       = "TCGA-SKCM",
  data.category = "DNA Methylation",
  data.type     = "Methylation Beta Value",
  platform      = "Illumina Human Methylation 450",
  data.format   = "Dataframe"
)

#GDCdownload(query_meth, method = "api", files.per.chunk = 10)
```

```{r}
methylation <- GDCprepare(query_meth)
```

```{r}
methylation_clinical <- as.data.frame(methylation@colData)
betas                <- as.data.frame(methylation@assays@data@listData)
cpg_sites            <- as.data.frame(methylation@rowRanges@elementMetadata)
```

```{r}
methylation_clinical <- methylation_clinical %>%
  mutate(
    stage_simple = gsub("Stage ", "", ajcc_pathologic_stage),
    stage_simple = gsub("[ABC]$", "", trimws(stage_simple))
  )

stage_mask <- methylation_clinical$stage_simple %in% c("II", "III") &
              !is.na(methylation_clinical$stage_simple)

methylation_clinical <- methylation_clinical[stage_mask, ]
betas_clean          <- betas[, stage_mask]
```

```{r}
site_mask <- !grepl("-", cpg_sites$gene) &
             !grepl(";", cpg_sites$gene) &
             !is.na(cpg_sites$gene) &
             complete.cases(betas_clean)

betas_clean <- betas_clean[site_mask, ]
cpg_sites   <- cpg_sites[site_mask, ]
```

```{r}
methylation_clinical$is_stageIII <- methylation_clinical$stage_simple == "III"

mval   <- t(apply(betas_clean, 1, function(x) log2(x / (1 - x))))
design <- model.matrix(~ is_stageIII, data = methylation_clinical)
fit    <- lmFit(mval, design)
fit2   <- eBayes(fit)
```

```{r}
# Chunk 8: Build Results Dataframe
dat_meth <- data.frame(
  foldchange = fit[["coefficients"]][, 2],
  logPvalue  = -log10(p.adjust(fit2[["p.value"]][, 2], method = "BH")),
  geneName   = cpg_sites$gene
)

dat_meth <- dat_meth %>%
  mutate(direction = case_when(
    foldchange >  1 & logPvalue > 2 ~ "Hypermethylated in III",
    foldchange < -1 & logPvalue > 2 ~ "Hypomethylated in III",
    TRUE ~ "Not significant"
  ))
```


```{r}
# Chunk 9+10: Volcano Plot with Significant Sites + Transcriptomic Genes Highlighted
sig_transcriptomic_genes <- dat %>%
  filter(direction != "Not significant") %>%
  pull(geneName)

dat_meth <- dat_meth %>%
  mutate(highlight = case_when(
    geneName %in% sig_transcriptomic_genes & foldchange >  1 & logPvalue > 2 ~ "Hypermethylated + DE in III",
    geneName %in% sig_transcriptomic_genes & foldchange < -1 & logPvalue > 2 ~ "Hypomethylated + DE in III",
    geneName %in% sig_transcriptomic_genes ~ "DE gene (not diff methylated)",
    direction == "Hypermethylated in III" ~ "Hypermethylated in III",
    direction == "Hypomethylated in III"  ~ "Hypomethylated in III",
    TRUE ~ "Not significant"
  ))

highlight_genes <- dat_meth %>% filter(highlight != "Not significant")

cols_combined <- c(
  "Hypermethylated + DE in III"   = "darkred",
  "Hypomethylated + DE in III"    = "darkblue",
  "DE gene (not diff methylated)" = "purple",
  "Hypermethylated in III"        = "firebrick",
  "Hypomethylated in III"         = "steelblue",
  "Not significant"               = "grey70"
)

ggplot(data = dat_meth, aes(x = foldchange, y = logPvalue)) +
  # Grey background dots first
  geom_point(data = dat_meth %>% filter(highlight == "Not significant"),
             aes(colour = highlight), alpha = 0.2, size = 0.6) +
  # Highlighted genes on top
  geom_point(data = dat_meth %>% filter(highlight != "Not significant"),
             aes(colour = highlight), alpha = 0.9, size = 2) +
  scale_colour_manual(values = cols_combined) +
  geom_vline(xintercept =  1, colour = "darkred", linetype = "dashed") +
  geom_vline(xintercept = -1, colour = "darkred", linetype = "dashed") +
  geom_hline(yintercept =  2, colour = "darkred", linetype = "dashed") +
  xlab("Fold Change (M-value)") +
  ylab("-log10 p value") +
  ggtitle("Methylation Volcano Plot: Stage III vs Stage II (TCGA-SKCM)\nwith Significant Sites and Transcriptomic DE Genes Highlighted") +
  theme_bw() +
  theme(legend.position = "top")
```


```{r}
# Install if needed
BiocManager::install("IlluminaHumanMethylation450kanno.ilmn12.hg19")

library(IlluminaHumanMethylation450kanno.ilmn12.hg19)

# Get full annotation
anno <- getAnnotation(IlluminaHumanMethylation450kanno.ilmn12.hg19)

# Look up cg17618595 specifically
anno["cg17618595", c("Name", "UCSC_RefGene_Name", "UCSC_RefGene_Group", 
                      "Relation_to_Island", "chr", "pos")]
```



another transcriptino volcano plot


```{r}
library(TCGAbiolinks)
library(limma)
library(ggplot2)
library(SummarizedExperiment)
library(dplyr)

query_rna <- GDCquery(
  project = "TCGA-SKCM",
  data.category = "Transcriptome Profiling",
  data.type = "Gene Expression Quantification",
  workflow.type = "STAR - Counts"
)
# GDCdownload(query_rna)
rna_se <- GDCprepare(query_rna)

counts_mat <- assay(rna_se, "unstranded")

sample_meta <- as.data.frame(colData(rna_se)) %>%
  mutate(
    barcode = rownames(.),
    stage_simple = gsub("Stage ", "", ajcc_pathologic_stage),
    stage_simple = gsub("[ABC]$", "", trimws(stage_simple))
  ) %>%
  filter(stage_simple %in% c("I", "II"),
         !is.na(stage_simple))

counts_sub <- counts_mat[, sample_meta$barcode]

keep <- rowSums(counts_sub >= 10) >= 5
counts_sub <- counts_sub[keep, ]
mval <- t(apply(counts_sub, 1, function(x) log2(x + 1)))

sample_meta$is_stageII <- sample_meta$stage_simple == "II"

design <- model.matrix(~ is_stageII, data = sample_meta)
fit    <- lmFit(mval, design)
fit2   <- eBayes(fit)

dat <- data.frame(
  foldchange = fit[["coefficients"]][, 2],
  logPvalue  = -log10(p.adjust(fit2[["p.value"]][, 2], method = "BH")),
  geneName   = rownames(counts_sub)
)
dat$threshold <- as.factor(abs(dat$foldchange) < 1)

library(org.Hs.eg.db)

ensembl_ids <- gsub("\\..*", "", dat$geneName)

gene_symbols <- mapIds(org.Hs.eg.db,
                       keys      = ensembl_ids,
                       column    = "SYMBOL",
                       keytype   = "ENSEMBL",
                       multiVals = "first")

dat$geneName <- ifelse(is.na(gene_symbols), ensembl_ids, gene_symbols)

library(ggrepel)

dat <- dat %>%
  mutate(direction = case_when(
    foldchange >  1 & logPvalue > 2 ~ "Upregulated in II",
    foldchange < -1 & logPvalue > 2 ~ "Downregulated in II",
    TRUE ~ "Not significant"
  ))

sig_genes <- dat %>% filter(direction != "Not significant")

force_label <- dat %>%
  filter(geneName %in% c("FOXQ1"))

cols <- c("Upregulated in II"   = "firebrick",
          "Downregulated in II" = "steelblue",
          "Not significant"     = "grey70")

ggplot(data = dat, aes(x = foldchange, y = logPvalue)) +
  geom_point(data = dat %>% filter(direction == "Not significant"),
             aes(colour = direction), alpha = 0.2, size = 0.6) +
  geom_point(data = dat %>% filter(direction != "Not significant"),
             aes(colour = direction), alpha = 0.9, size = 1.5) +
  scale_colour_manual(values = cols) +
  geom_text_repel(
    data = force_label,
    aes(label = geneName),
    size = 3,
    max.overlaps = 20,
    box.padding = 0.5,
    fontface = "bold"
  ) +
  geom_vline(xintercept =  1, colour = "darkred", linetype = "dashed") +
  geom_vline(xintercept = -1, colour = "darkred", linetype = "dashed") +
  geom_hline(yintercept =  2, colour = "darkred", linetype = "dashed") +
  xlab("Fold Change") +
  ylab("-log10 p value") +
  ggtitle("Volcano Plot: Stage II vs Stage I (TCGA-SKCM)") +
  theme_bw() +
  theme(legend.position = "top")

```
YES KEEP - methylation volcano plot stages 1 to 2
new volcano methylation plot
```{r}
library(TCGAbiolinks)
library(limma)
library(ggplot2)
library(ggrepel)
library(dplyr)
```

```{r}
if (!requireNamespace("remotes", quietly = TRUE)) {
  install.packages("remotes")
}
library("remotes")
remotes::install_github("BioinformaticsFMRP/TCGAbiolinks")
```

```{r}
#query_meth <- GDCquery(
  #project       = "TCGA-SKCM",
  #data.category = "DNA Methylation",
  #data.type     = "Methylation Beta Value",
  #platform      = "Illumina Human Methylation 450",
  #data.format   = "Dataframe"
#)

#GDCdownload(query_meth, method = "api", files.per.chunk = 10)
```

```{r}
#methylation <- GDCprepare(query_meth)
```

```{r}
methylation_clinical <- as.data.frame(methylation@colData)
betas                <- as.data.frame(methylation@assays@data@listData)
cpg_sites            <- as.data.frame(methylation@rowRanges@elementMetadata)
```

```{r}
methylation_clinical <- methylation_clinical %>%
  mutate(
    stage_simple = gsub("Stage ", "", ajcc_pathologic_stage),
    stage_simple = gsub("[ABC]$", "", trimws(stage_simple))
  )

stage_mask <- methylation_clinical$stage_simple %in% c("I", "II") &
              !is.na(methylation_clinical$stage_simple)

methylation_clinical <- methylation_clinical[stage_mask, ]
betas_clean          <- betas[, stage_mask]
```

```{r}
site_mask <- !grepl("-", cpg_sites$gene) &
             !grepl(";", cpg_sites$gene) &
             !is.na(cpg_sites$gene) &
             complete.cases(betas_clean)

betas_clean <- betas_clean[site_mask, ]
cpg_sites   <- cpg_sites[site_mask, ]
```

```{r}
methylation_clinical$is_stageII <- methylation_clinical$stage_simple == "II"

mval   <- t(apply(betas_clean, 1, function(x) log2(x / (1 - x))))
design <- model.matrix(~ is_stageII, data = methylation_clinical)
fit    <- lmFit(mval, design)
fit2   <- eBayes(fit)
```

```{r}
# Chunk 8: Build Results Dataframe - Stage II vs I
dat_meth <- data.frame(
  foldchange = fit[["coefficients"]][, 2],
  logPvalue  = -log10(p.adjust(fit2[["p.value"]][, 2], method = "BH")),
  geneName   = cpg_sites$gene
)

dat_meth <- dat_meth %>%
  mutate(direction = case_when(
    foldchange >  1 & logPvalue > 2 ~ "Hypermethylated in II",
    foldchange < -1 & logPvalue > 2 ~ "Hypomethylated in II",
    TRUE ~ "Not significant"
  ))

```

```{r}
# Chunk 9+10: Volcano Plot with Significant Sites Labeled + Transcriptomic Genes Highlighted
sig_transcriptomic_genes <- dat %>%
  filter(direction != "Not significant") %>%
  pull(geneName)

dat_meth <- dat_meth %>%
  mutate(highlight = case_when(
    geneName %in% sig_transcriptomic_genes & foldchange >  1 & logPvalue > 2 ~ "Hypermethylated + DE in II",
    geneName %in% sig_transcriptomic_genes & foldchange < -1 & logPvalue > 2 ~ "Hypomethylated + DE in II",
    geneName %in% sig_transcriptomic_genes ~ "DE gene (not diff methylated)",
    direction == "Hypermethylated in II" ~ "Hypermethylated in II",
    direction == "Hypomethylated in II"  ~ "Hypomethylated in II",
    TRUE ~ "Not significant"
  ))

highlight_genes <- dat_meth %>% filter(highlight != "Not significant")

cols_combined <- c(
  "Hypermethylated + DE in II"    = "darkred",
  "Hypomethylated + DE in II"     = "darkblue",
  "DE gene (not diff methylated)" = "purple",
  "Hypermethylated in II"         = "firebrick",
  "Hypomethylated in II"          = "steelblue",
  "Not significant"               = "grey70"
)

ggplot(data = dat_meth, aes(x = foldchange, y = logPvalue)) +
  # Grey background dots
  geom_point(data = dat_meth %>% filter(highlight == "Not significant"),
             aes(colour = highlight), alpha = 0.2, size = 0.6) +
  # Purple DE genes
  geom_point(data = dat_meth %>% filter(highlight == "DE gene (not diff methylated)"),
             aes(colour = highlight), alpha = 0.9, size = 2) +
  # Significant methylation sites - bigger and on top
  geom_point(data = dat_meth %>% filter(highlight %in% c("Hypomethylated in II",
                                                          "Hypermethylated in II",
                                                          "Hypomethylated + DE in II",
                                                          "Hypermethylated + DE in II")),
             aes(colour = highlight), alpha = 1, size = 4) +
  scale_colour_manual(values = cols_combined) +
  geom_vline(xintercept =  1, colour = "darkred", linetype = "dashed") +
  geom_vline(xintercept = -1, colour = "darkred", linetype = "dashed") +
  geom_hline(yintercept =  2, colour = "darkred", linetype = "dashed") +
  xlab("Fold Change (M-value)") +
  ylab("-log10 p value") +
  ggtitle("Methylation Volcano Plot: Stage II vs Stage I (TCGA-SKCM)\nwith Significant Sites and Transcriptomic DE Genes Highlighted") +
  theme_bw() +
  theme(legend.position = "top")
```

YES KEEP - stages 2 to 3 spearman correlation bar graph 
```{r}
library(ggplot2)
library(dplyr)
library(org.Hs.eg.db)

sample_meta_23 <- as.data.frame(colData(rna_se)) %>%
  mutate(
    barcode = rownames(.),
    stage_simple = gsub("Stage ", "", ajcc_pathologic_stage),
    stage_simple = gsub("[ABC]$", "", trimws(stage_simple))
  ) %>%
  filter(stage_simple %in% c("II", "III"),
         !is.na(stage_simple))

counts_sub_23 <- counts_mat[, sample_meta_23$barcode]
keep_23 <- rowSums(counts_sub_23 >= 10) >= 5
counts_sub_23 <- counts_sub_23[keep_23, ]

ensembl_ids_23 <- gsub("\\..*", "", rownames(counts_sub_23))
gene_symbols_23 <- mapIds(org.Hs.eg.db,
                          keys      = ensembl_ids_23,
                          column    = "SYMBOL",
                          keytype   = "ENSEMBL",
                          multiVals = "first")

foxq1_idx_23 <- which(gene_symbols_23 == "FOXQ1")
foxq1_counts_23 <- as.numeric(counts_sub_23[foxq1_idx_23[1], ])

chromatin_genes <- c(
  "SUZ12", "BMI1", "CBX7", "EED", "RNF2", "AEBP2",  # Polycomb
  "DNMT3B",                                            # DNA Methylation
  "KDM6A", "KDM5C",                                   # Histone Demethylase
  "SETDB1",                                            # H3K9 Methyltransferase
  "SIRT1", "HDAC3",                                   # Histone Deacetylase
  "SMARCC1", "SETD2"                                  # Other
)

cor_results <- data.frame(gene = character(), r = numeric(),
                          pval = numeric(), stringsAsFactors = FALSE)

for (gene in chromatin_genes) {
  gene_idx <- which(gene_symbols_23 == gene)
  if (length(gene_idx) > 0) {
    gene_counts <- as.numeric(counts_sub_23[gene_idx[1], ])
    if (length(foxq1_counts_23) == length(gene_counts)) {
      cor_test <- cor.test(foxq1_counts_23, gene_counts, method = "spearman")
      cor_results <- rbind(cor_results,
                           data.frame(gene = gene,
                                      r    = round(cor_test$estimate, 4),
                                      pval = cor_test$p.value,
                                      stringsAsFactors = FALSE))
    }
  }
}

# ── 5. Keep only significant results and add direction ────────────────────────
cor_results <- cor_results %>%
  filter(pval < 0.05) %>%
  mutate(
    direction = case_when(
      r < 0 ~ "Negative (potential FOXQ1 silencer)",
      r > 0 ~ "Positive (co-expressed with FOXQ1)"
    ),
    sig_label = case_when(
      pval < 0.001 ~ "***",
      pval < 0.01  ~ "**",
      pval < 0.05  ~ "*",
      TRUE         ~ ""
    )
  ) %>%
  arrange(r)

cor_results$gene <- factor(cor_results$gene, levels = cor_results$gene)

# ── 6. Plot ───────────────────────────────────────────────────────────────────
ggplot(cor_results, aes(x = r, y = gene, fill = direction)) +
  geom_bar(stat = "identity", width = 0.7) +
  geom_text(aes(label = sig_label,
                x = ifelse(r >= 0, r + 0.004, r - 0.004)),
            size = 4, vjust = 0.75) +
  scale_fill_manual(values = c(
    "Negative (potential FOXQ1 silencer)" = "firebrick",
    "Positive (co-expressed with FOXQ1)"  = "steelblue"
  ), name = "Direction") +
  geom_vline(xintercept = 0, color = "black", linewidth = 0.8) +
  xlab("Spearman r") +
  ylab("") +
  ggtitle("Spearman Correlation: Chromatin Regulators vs FOXQ1",
          subtitle = "Stage II vs III (TCGA-SKCM) — * p<0.05  ** p<0.01  *** p<0.001") +
  theme_bw() +
  theme(
    legend.position    = "right",
    plot.subtitle      = element_text(size = 9),
    axis.text.y        = element_text(size = 11, face = "bold"),
    panel.grid.major.y = element_line(color = "grey90")
  )
```
stage 1 to 2:


spearman for genes that upregulate from 1 to 2:

```{r}
# Recreate Stage I vs II objects
sample_meta_12 <- as.data.frame(colData(rna_se)) %>%
  mutate(
    barcode = rownames(.),
    stage_simple = gsub("Stage ", "", ajcc_pathologic_stage),
    stage_simple = gsub("[ABC]$", "", trimws(stage_simple))
  ) %>%
  filter(stage_simple %in% c("I", "II"),
         !is.na(stage_simple))

counts_sub_12 <- counts_mat[, sample_meta_12$barcode]
keep_12 <- rowSums(counts_sub_12 >= 10) >= 5
counts_sub_12 <- counts_sub_12[keep_12, ]

library(org.Hs.eg.db)
ensembl_ids_12 <- gsub("\\..*", "", rownames(counts_sub_12))
gene_symbols_12 <- mapIds(org.Hs.eg.db,
                          keys      = ensembl_ids_12,
                          column    = "SYMBOL",
                          keytype   = "ENSEMBL",
                          multiVals = "first")

# Recreate FOXQ1 counts for Stage I vs II
foxq1_idx_12 <- which(gene_symbols_12 == "FOXQ1")
foxq1_counts_12 <- as.numeric(counts_sub_12[foxq1_idx_12[1], ])

activation_genes <- c("MITF", "SOX10", "PAX3", "TFAP2A",
                      "YAP1", "TEAD1", "RUNX2", "SP1", "MYC",
                      "CTNNB1", "WNT5A", "LEF1", "TCF7", "AXIN2",
                      "EGFR", "MET", "IGF1R", "AKT1", "MAPK1",
                      "TP53", "RB1", "PTEN", "CDKN2A", "KLF4")

cor_activation <- data.frame(gene = character(), r = numeric(),
                             pval = numeric(), stringsAsFactors = FALSE)

for (gene in activation_genes) {
  gene_idx <- which(gene_symbols_12 == gene)
  if (length(gene_idx) > 0) {
    gene_counts <- as.numeric(counts_sub_12[gene_idx[1], ])
    if (length(foxq1_counts_12) == length(gene_counts)) {
      cor_test <- cor.test(foxq1_counts_12, gene_counts, method = "spearman")
      cor_activation <- rbind(cor_activation,
                              data.frame(gene = gene,
                                         r    = round(cor_test$estimate, 4),
                                         pval = cor_test$p.value,
                                         stringsAsFactors = FALSE))
    }
  }
}


```
YES KEEP - spearman correlation
```{r}
cor_activation_sig <- cor_activation %>%
  filter(pval < 0.05, r > 0) %>%  
  mutate(
    direction = "Positive (potential FOXQ1 activator)",
    sig_label = case_when(
      pval < 0.001 ~ "***",
      pval < 0.01  ~ "**",
      pval < 0.05  ~ "*",
      TRUE         ~ ""
    )
  ) %>%
  arrange(r)

cor_activation_sig$gene <- factor(cor_activation_sig$gene, 
                                   levels = cor_activation_sig$gene)

ggplot(cor_activation_sig, aes(x = r, y = gene, fill = direction)) +
  geom_bar(stat = "identity", width = 0.7) +
  geom_text(aes(label = sig_label, x = r + 0.01),
            size = 4, vjust = 0.75) +
  scale_fill_manual(values = c(
    "Positive (potential FOXQ1 activator)" = "steelblue"
  ), name = "Direction") +
  geom_vline(xintercept = 0, color = "black", linewidth = 0.8) +
  xlab("Spearman r") +
  ylab("") +
  ggtitle("Spearman Correlation: Signaling Genes vs FOXQ1",
          subtitle = "Stage I vs II (TCGA-SKCM) — * p<0.05  ** p<0.01  *** p<0.001") +
  theme_bw() +
  theme(
    legend.position    = "right",
    plot.subtitle      = element_text(size = 9),
    axis.text.y        = element_text(size = 11, face = "bold"),
    panel.grid.major.y = element_line(color = "grey90")
  )
```

OTHER GRAPHS 

Lollipops

```{r}
library(maftools)

maf_query <- GDCquery(
  project = 'TCGA-SKCM',
  data.category = "Simple Nucleotide Variation",
  access = "open",
  data.type = "Masked Somatic Mutation",
  workflow.type = "Aliquot Ensemble Somatic Variant Merging and Masking")

# GDCdownload(maf_query)  # already downloaded
maf <- GDCprepare(maf_query)

clinic$Tumor_Sample_Barcode <- clinic$bcr_patient_barcode

maf_object <- read.maf(maf = maf, clinicalData = clinic, isTCGA = TRUE)

stage_I_barcodes <- clinic$Tumor_Sample_Barcode[
  grepl("Stage IA|Stage IB|Stage I$",
        clinic$ajcc_pathologic_tumor_stage, ignore.case = TRUE)
]

stage_II_barcodes <- clinic$Tumor_Sample_Barcode[
  grepl("Stage IIA|Stage IIB|Stage IIC|Stage II$",
        clinic$ajcc_pathologic_tumor_stage, ignore.case = TRUE)
]

stage_III_barcodes <- clinic$Tumor_Sample_Barcode[
  grepl("Stage IIIA|Stage IIIB|Stage IIIC|Stage III$",
        clinic$ajcc_pathologic_tumor_stage, ignore.case = TRUE)
]

maf_stage_I   <- subsetMaf(maf = maf_object, tsb = stage_I_barcodes)
maf_stage_II  <- subsetMaf(maf = maf_object, tsb = stage_II_barcodes)
maf_stage_III <- subsetMaf(maf = maf_object, tsb = stage_III_barcodes)

lollipopPlot2(m1 = maf_stage_I,
              m2 = maf_stage_II,
              gene = "FOXQ1",
              m1_name = "Stage I",
              m2_name = "Stage II")

lollipopPlot2(m1 = maf_stage_II,
              m2 = maf_stage_III,
              gene = "FOXQ1",
              m1_name = "Stage II",
              m2_name = "Stage III")
```
KM plots

```{r}

```


