---
title: "Final_Project_Code"
output: html_document
---
```{r setup}
 #knitr::opts_knit$set(root.dir = normalizePath("/home1/gittings/490_cluster/analysis_data")) 
#knitr::opts_knit$set(root.dir = normalizePath("/home1/emmashih/490_cluster/analysis_data")) 
       knitr::opts_knit$set(root.dir = normalizePath("/home1/spoden/490_cluster/analysis_data")) 
```

```{r}
library(TCGAbiolinks)
library(maftools)
library(ggplot2)
```

```{r}
# code to download and load CLINICAL DATA
clin_query <- GDCquery(project = "TCGA-SKCM",
data.category = "Clinical",
data.type = "Clinical Supplement",
data.format = 'SKCM Biotab')
# GDCdownload(clin_query)
clinical.BCRtab.all <- GDCprepare(clin_query)
clinic <- clinical.BCRtab.all$clinical_patient_skcm[-c(1,2),]
```

```{r}
# code to download METHYLATION DATA (code to load data below this chunk)
library(BiocManager)
library(maftools)
library(TCGAbiolinks)
library(ggplot2)

clinic$Tumor_Sample_Barcode <- clinic$bcr_patient_barcode

maf_query <- GDCquery(
  project = 'TCGA-SKCM',
  data.category = "Simple Nucleotide Variation",
  access = "open",
  data.type = "Masked Somatic Mutation",
  workflow.type = "Aliquot Ensemble Somatic Variant Merging and Masking"
)

# GDCdownload(maf_query)
setwd("/home1/emmashih/490_cluster/analysis_data")
maf <- GDCprepare(maf_query)
maf_object <- read.maf(maf = maf, clinicalData = clinical, isTCGA = TRUE)


# ill resave the clinical dataframe and maf object from today
write.csv(x = clinical,
          file = "skcm_clinical_data.csv", row.names = F)

saveRDS(maf_object,
        file = "skcm_maf_object.RDS")
```

```{r}
# code to LOAD data
# load clinical data
clinical <- read.csv("/home1/emmashih/490_cluster/analysis_data/skcm_clinical_data.csv")

# load maf object
maf_object <- readRDS("/home1/emmashih/490_cluster/analysis_data/skcm_maf_object.RDS")
```



```{r Exercise 4.2}
clinic$stage_simple <- ifelse(grepl("^Stage IV", clinic$ajcc_pathologic_tumor_stage), 4,
                           ifelse(grepl("^Stage III", clinic$ajcc_pathologic_tumor_stage), 3,
                           ifelse(grepl("^Stage II", clinic$ajcc_pathologic_tumor_stage), 2,
                           ifelse(grepl("^Stage I($|[^VIX])", clinic$ajcc_pathologic_tumor_stage), 1, NA))))

clinic$age <- as.numeric(clinic$age_at_diagnosis)

clinic_scatter <- clinic[!is.na(clinic$stage_simple) & !is.na(clinic$age), ]

plot(jitter(clinic_scatter$stage_simple, amount = 0.1), clinic_scatter$age,
     xlab = "Tumor Stage",
     ylab = "Age at Diagnosis",
     main = "Scatterplot of Age vs Tumor Stage",
     pch = 19)
```{r}
ggplot(clinic, aes(x = ajcc_pathologic_tumor_stage)) + geom_bar()
```

```{r}
library(dplyr)
library(survival)
library(survminer)

colnames(clinic)
```
```{r}

clinic <- clinic %>%
  mutate(
    death_days_to        = as.numeric(death_days_to),
    last_contact_days_to = as.numeric(last_contact_days_to),
    age_at_diagnosis     = as.numeric(age_at_diagnosis),
    breslow_thickness_at_diagnosis = as.numeric(breslow_thickness_at_diagnosis),
    
    # Unified survival time: use death days if dead, otherwise last contact
    os_time  = ifelse(!is.na(death_days_to), death_days_to, last_contact_days_to),
    os_event = ifelse(vital_status == "Dead", 1, 0)
  )
```
```{r}
fit <- survfit(Surv(os_time, os_event) ~ 1, data = clinic)

ggsurvplot(fit, data = clinic,
           #risk.table = TRUE,
           conf.int   = TRUE,
           title      = "TCGA SKCM: Overall Survival",
           xlab       = "Days",
           ylab       = "Survival Probability")
```


random plot

```{r}
library(TCGAbiolinks)
library(maftools)
```


boxplot of Breslow thickness across AJCC tumor stages (the one from the exit ticket but better verison)
```{r}
library(ggplot2)
library(dplyr)

# Clean the stage and Breslow columns
breslow_clean <- clinic %>%
  filter(
    !is.na(ajcc_pathologic_tumor_stage),
    !is.na(breslow_thickness_at_diagnosis),
    breslow_thickness_at_diagnosis > 0,
    !ajcc_pathologic_tumor_stage %in% c("[Not Available]", "[Discrepancy]", 
                                         "[Unknown]", ""),
    !breslow_thickness_at_diagnosis %in% c("[Not Available]", "[Unknown]", "")
  ) %>%
  mutate(
    breslow_thickness_at_diagnosis = as.numeric(breslow_thickness_at_diagnosis),
    # Simplify stage to I, II, III, IV (strip substage letters like IA, IIB, etc.)
    stage_simple = gsub("Stage ", "", ajcc_pathologic_tumor_stage),
    stage_simple = gsub("[ABC]$", "", stage_simple),
    stage_simple = factor(stage_simple, levels = c("I", "II", "III", "IV"))
  ) %>%
  filter(!is.na(stage_simple), !is.na(breslow_thickness_at_diagnosis))

ggplot(breslow_clean, aes(x = stage_simple, y = breslow_thickness_at_diagnosis, 
                           fill = stage_simple)) +
  geom_boxplot(outlier.shape = 21, outlier.alpha = 0.5) +
  scale_fill_manual(values = c("I" = "blue", "II" = "red", 
                                "III" = "pink", "IV" = "green")) +
  scale_y_log10() +   # log scale because Breslow thickness is right-skewed
  xlab("AJCC Tumor Stage") +
  ylab("Breslow Thickness at Diagnosis (mm, log scale)") +
  ggtitle("Tumor Invasion Depth Across SKCM Stages")
```

Patient count by stage and vital status

```{r}
library(ggplot2)
library(dplyr)

# Data preparation
stage_survival <- clinic %>%
  filter(
    !is.na(ajcc_pathologic_tumor_stage),
    !ajcc_pathologic_tumor_stage %in% c("[Not Available]", "[Discrepancy]", 
                                         "[Unknown]", ""),
    !is.na(vital_status),
    !vital_status %in% c("[Not Available]", "[Unknown]", "")
  ) %>%
  mutate(
    stage_simple = gsub("Stage ", "", ajcc_pathologic_tumor_stage),
    stage_simple = gsub("[ABC]$", "", stage_simple),
    stage_simple = trimws(stage_simple),
    stage_simple = factor(stage_simple, levels = c("I", "II", "III", "IV")),
    vital_status = factor(vital_status, levels = c("Alive", "Dead"))
  ) %>%
  filter(!is.na(stage_simple))

# Calculate percentages
stage_pct <- stage_survival %>%
  count(stage_simple, vital_status) %>%
  group_by(stage_simple) %>%
  mutate(
    pct = n / sum(n) * 100,
    label = paste0(round(pct, 1), "%")
  )

# Plot
ggplot(stage_pct, aes(x = stage_simple, y = n, fill = vital_status)) +
  geom_bar(stat = "identity") +
  geom_text(
    aes(label = label),
    position = position_stack(vjust = 0.5),
    size = 3.5,
    color = "black",
    fontface = "bold"
  ) +
  scale_fill_manual(values = c("Alive" = "pink", "Dead" = "seagreen"),
                    name = "Vital Status") +
  labs(
    title = "SKCM Patient Count and Survival Outcome by Tumor Stage",
    x = "AJCC Tumor Stage",
    y = "Number of Patients"
  ) +
  theme_gray()
```





